---
title: "STATS 547 Homework 4"
author: "Ben Tward"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

My proposed method revolves around the idea of padding.

Let $k=\text{max}\{|e| : e\in\mathcal{E}\}$, the maximum cardinality of the set of edges. Then, for every $e$ in $\mathcal{E}$ with $|e|<k$, we add artificial nodes constructed using an averaging technique on the nodes within that edge. This will create a $k$-uniform hypergraph, and we can now construct our supersymmetric adjascency tensor $\text{A} \in \mathbb{R}^{\overbrace{n \times \dots \times n}^{k}}$.

The main advantage of this method is that .... i dont even know.

A disadvantage is that is sucks.

## Question 2

```{r, include=FALSE}
path = "C:\\Users\\Ben\\OneDrive\\Documents\\GitHub\\STATS-547\\ProblemSet4\\mouse_neurons.csv"
mouse_neurons = read.csv(path)
```

```{r}
multicor = function(data, method) {
  R = cor(data)
  k = ncol(R)
  eigenvals = eigen(R)$values  
  
  if (method == 1) {
    r = 1 - eigenvals[length(eigenvals)]
  } else if (method ==2) {
    r = (1 - det(R))^0.5
  } else if (method == 3) {
    r = sd(eigenvals) / sqrt(k)
  }
  
  return(r)

}
```

Gonna switch to Python so I can use HAT. Except I cant get the package to load.

```{python}

```
